A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-26 paper/fascd.tex
draft abstract, and clean intro

\begin{abstract} FIXME
\end{abstract}

@@

The constraint decomposition (CD) methods of Tai \cite{Tai2003} are designed to solve coercive variational inequality (VI) problems which arise from minimization of a convex functional over a convex set.  These methods have been shown to converge for multilevel and domain decompositions in a finite element (FE) context.  The multilevel CD method has been shown to have almost grid-independent error bounds, and near-optimal complexity, for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}; see also \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.  In this paper we extend the multilevel CD method by replacing the coarse-level corrections with a full approximation storage (FAS; \cite{Brandt1977,Bruneetal2015}) approach, and we apply this new solver to nonlinear VI problems which do not arise from minimization.

Specifically, our \emph{full approximation storage constraint decomposition} (FASCD) method extends the algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} in the following directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems.  The needed computations do not require an objective function, nor do they require linearity of the residual operator.  Section \ref{sec:results} shows results for several nonlinear problems.
\item By using $O(N)$-per-application Newton-type active set methods as smoothers, with a fixed number of Krylov iterations per step, we demonstrate mesh-independent convergence rates substantially better than in the current literature.
\item We implement both V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (also known as full multigrid; Algorithm \ref{alg:fascd-fcycle}).
\item Box constraints, namely upper and lower obstacles, are permitted; see Section \ref{sec:femultilevel}.
\item We observe an asymmetry special to multilevel CD methods, namely that up-smoothing is more effective than down-smoothing because a ``telescoping'' subset property can be exploited on the up-smoothing side.
\end{enumerate}

FASCD exhibits essentially the same optimal, mesh-independent convergence rates for VI problems that FAS multigrid exhibits for the corresponding unconstrained PDE problems.  The results in Section \ref{sec:results} include textbook multigrid efficiency, namely solutions within discretization error in work comparable to ten smoother sweeps on the finest-level mesh \cite{BrandtLivne2011}, of F-cycles applied to classical Laplacian obstacle problems (Example \ref{ex:results:classical}), $p$-Laplacian obstacle problems (Example \ref{ex:results:plap}), and nonnegativity-constrained advection-diffusion equations (Example \ref{ex:results:advdiff}).  Furthermore, these results are essentially insensitive to the length or structure of the free boundary, i.e.~the geometric complexity of the coincidence set (Examples \ref{ex:results:classical}, \ref{ex:results:advdiff}).

The iterates from FASCD are admissible, and thus the nonlinear operator in the VI needs only be defined for admissible states.  For example, we solve an ice sheet geometry problem (Examples \ref{ex:sia}, \ref{ex:results:sia}) in which the ice velocity is only meaningful for surface elevation iterates which exceed the bed elevation.  More generally, fluid-layer problems, with a non-negative thickness constraint, would require the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation \cite{Bueler2021conservation}.  (Though a target for FASCD, such problems, however, are not addressed here.)  Note that non-admissible methods, including semi-smooth methods \cite{BensonMunson2006}, would require unnatural modifications of the operator formula in such a fluid-layer context.  In any case, FASCD solves the difficult ice sheet problem, wherein the solution has unbounded gradient at the free boundary, by a mesh-independent number of iterations, with near textbook efficiency.

For the classical obstacle problem, based on the linear, symmetric Laplacian operator, certain multilevel techniques, such as the truncated monotone multigrid method \cite{GraeserKornhuber2009,Kornhuber1994}, are known to improve performance relative to the original multilevel CD method \cite{Tai2003}.  These improved methods modify the nodal basis functions.  By contrast FASCD makes essentially no assumptions about the FE discretization.  Acceleration via basis-level manipulations is a topic for future research.

This paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive variational inequalities, with several example problems, and then Section \ref{sec:cd} reviews the basics of CDs, including the associated iterations, as proposed by Tai \cite{Tai2003}.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, along with the special monotone restriction operators used here.  Section \ref{sec:cdmultilevel} presents defect constraints, leading to an apparently new understanding of telescoping sets for up-smoothing.  Section \ref{sec:vcycle} then presents the FASCD V-cycle, with Section \ref{sec:implementation} addressing implementation and smoother semantics, plus F-cycles.  Section \ref{sec:results} gives FASCD numerical results for the Section \ref{sec:vi} problems, followed by discussion and outlook in Section \ref{sec:discussion}.  An Appendix sketchs how FASCD reduces to classical FAS multigrid for PDEs.

>>>>

\begin{abstract}
The \emph{full approximation storage constraint decomposition} (FASCD) multilevel finite element method introduced here is a common extension of the full approximation storage multigrid technique for nonlinear partial differential equations, due to A.~Brandt, and of the constraint decomposition method introduced by X.-C.~Tai for variational inequalities.  We extend the existing constraint decomposition ideas by recognizing and exploiting the telescoping nature of certain function space subset decompositions arising from multilevel mesh hierarchies.  Then, by using strong smoothers with work proportional to the number of unknowns on the finite element mesh level, we demonstrate that FASCD V-cycles show superior mesh-independent convergence rates and that FASCD F-cycles (full multigrid) are optimal solvers.  The example problems include unilateral and bi-obstacle variational inequality problems with linear and nonlinear differential operators.
\end{abstract}

@@

The constraint decomposition (CD) methods of Tai \cite{Tai2003} are designed to solve coercive variational inequality (VI) problems which arise from minimization of a convex functional over a convex set.  These methods have been shown to converge for multilevel and domain decompositions in a finite element (FE) context.  In the multilevel case, the CD method has been shown to have almost grid-independent error bounds, and near-optimal complexity, for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}; see also \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.  In this paper we extend the multilevel CD method by replacing the coarse-level corrections with a full approximation storage (FAS; \cite{Brandt1977,Bruneetal2015}) approach, and we apply this new solver to nonlinear VI problems which do not arise from minimization.

Specifically, our \emph{full approximation storage constraint decomposition} (FASCD) method extends the algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} in the following directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems.  The needed computations require neither an objective function nor linearity of the residual operator.
\item Box constraints, namely upper and lower obstacles, are permitted; see Section \ref{sec:femultilevel}.
\item We observe an asymmetry special to multilevel CD methods, namely that up-smoothing is more effective than down-smoothing because a ``telescoping'' subset property can be exploited on the up-smoothing side.
\item We implement V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle})
\item By using a Newton-type active set method as a smoother, with a fixed number of Krylov iterations per step so that the work of the smoother is proportional to the number of unknowns on the given level, we demonstrate that F-cycle FASCD is an optimal solver.
\end{enumerate}

FASCD V-cycles exhibit essentially the same mesh-independent convergence rates for VI problems that FAS multigrid V-cycles exhibit for the corresponding (unconstrained) PDE problems.  The F-cycle results in Section \ref{sec:results} show nearly-textbook multigrid efficiency, namely solutions within discretization error in work comparable to few smoother sweeps on the finest-level mesh \cite{BrandtLivne2011}, for classical Laplacian obstacle problems, $p$-Laplacian obstacle problems, and nonnegativity-constrained advection-diffusion equations.  Furthermore, these results are essentially insensitive to the length or structure of the free boundary, i.e.~the geometric complexity of the coincidence set.

The iterates from FASCD are admissible, and thus the nonlinear operator in the VI needs only be defined for admissible states.  For example, we solve an ice sheet geometry problem (Examples \ref{ex:sia}, \ref{ex:results:sia}) in which the ice flow formulas are only meaningful for surface elevation iterates which exceed the bed elevation.  (More generally, fluid-layer problems, with a non-negative thickness constraint, often require the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation \cite{Bueler2021conservation}.  However, though a target for future applications of FASCD, such problems are not addressed here.)  Note that non-admissible methods, including semi-smooth methods \cite{BensonMunson2006}, would require unnatural modifications of the operator formula in such a fluid-layer context.  FASCD solves this low-regularity ice sheet problem, wherein the solution has unbounded gradient at the free boundary, by a mesh-independent number of F-cycle iterations.

This paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, with several example problems.  Then Section \ref{sec:cd} reviews the basics of CDs, including the associated iterations, as proposed by Tai \cite{Tai2003}.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, along with special monotone restriction operators \cite{GraeserKornhuber2009}.  Section \ref{sec:cdmultilevel} presents defect constraints \cite{GraeserKornhuber2009}, leading to an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD.  Section \ref{sec:vcycle} then states the FASCD V-cycle, with Section \ref{sec:implementation} addressing implementation concerns, smoother semantics, and F-cycles.  Section \ref{sec:results} gives FASCD numerical results, for problems in Section \ref{sec:vi}, from a parallel Firedrake \cite{Rathgeberetal2016} implementation.  Discussion and outlook in Section \ref{sec:discussion} and an Appendix, which sketchs how FASCD reduces to classical FAS multigrid for PDEs, finish the work.
